---
title: "PHP2511_HW1"
author: "Kuan-Min Lee"
date: "2023-02-11"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 1 Simple Linear Regression

This Rmd file is created for Assignment 1 from PHP2511 course taught by Dr. Muillio in Spring 2023 semester at Brown University.

## Exploratory Analysis Portion

In this portion of the assignment, the dataset from paper "Association of Highly Restrictive State Abortion Policies With Abortion Rates, 2000-2014" is utilized, and the exploratory data analysis will be performed on this dataset.

To start with this, we begin with loading the excel file of the dataset and viewing out the data structure of the dataset.

We start with loading all libraries that are necessary for loading the dataset:
```{r setup, include=FALSE}
library(readxl) # import library for reading excel file
```

Next, we read in the excel file of the dataset and view the dimension of the dataset:
```{r}
dataset_test <- read_excel("state_rep_laws.xlsx") # read in excel file for the dataset
```

```{r cars}
```{r}

```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
